Combinatorial Proofs and Decomposition Theorem for First-order Logic

### 2021-01-07 main.tex
Reorganize inference rules and figures

\documentclass[conference]{IEEEtran}
% Some Computer Society conferences also require the compsoc mode option,
% but others use the standard conference format.
%
% If IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it like:
% \documentclass[conference]{../sty/IEEEtran}

@@

\section{Introduction}

@@

\bfit{Terms} $s, t, u, \ldots$ and \bfit{atoms} (atomic formulas) $\alpha, \beta,
\gamma, \ldots$ are generated inductively from variables $x, y, z, \ldots$: if
$\delta$ (resp. $p$) is an $n$-ary function (resp. predicate) symbol and $t_1, \ldots,
t_n$ are terms, then $\delta t_1 \ldots t_n$ (resp. $pt_1 \ldots t_n$) is a term
(resp. atom). The set of atoms is extended with logical constant $\ttt$ and
$\fff$. We also assume that each predicate symbol $p$ has a \bfit{dual}
predicate symbol $\overline{p}$ with $\overline{p} \neq p$ and
$\overline{\overline{p}} = p$. We extend the duality to atoms with
$\overline{pt_1 \ldots t_n} = \overline{p}t_1 \ldots t_n$, $\overline{\ttt} =
\fff$, and $\overline{\fff} = \ttt$. \bfit{Formulas} $A, B, C$ are generated inductively from atoms by binary
connectives $\cor$ and $\cand$ and quantifiers $\forall x$ and $\exists x$.
The \bfit{negation} $\neg$ is defined in the usual way. A formula is
\bfit{rectified} if all bound variables are distinct from one another and from
all free variables. Without loss of generality, we assume that all formulas are
rectifed throughtout the paper as every formula has a logically equivalent
rectified form. A formula is \bfit{semi-rectified} if it can be rectified in a
unique way via variable renaming.

@@

A \bfit{sequent} is a multiset of formulas and sequents are proved using the
following inference rules.
\begin{center}

\mbox{
\begin{prooftree}
  \infer0[$\ax$]{\vdash A, \neg A}
\end{prooftree}
\hspace{2cm}

\begin{prooftree}
  \hypo{\vdash \Gamma, A}
  \hypo{\vdash \Delta, \neg A}
  \infer2[$\cut$]{\vdash \Gamma, \Delta}
\end{prooftree}}
\vspace{0.2cm}

\mbox{
\begin{prooftree}
  \hypo{\vdash \Gamma, A}
  \hypo{\vdash \Delta, B}
  \infer2[$\cand$]{\vdash \Gamma, \Delta, A \cand B}
\end{prooftree}
\hspace{1.5cm}

\begin{prooftree}
  \hypo{\vdash \Gamma, A, B}
  \infer1[$\cor$]{\vdash \Gamma, A \cor B}
\end{prooftree}}
\vspace{0.2cm}

\mbox{
\begin{prooftree}
  \hypo{\vdash \Gamma, A, A}
  \infer1[$\ctr$]{\vdash \Gamma, A}
\end{prooftree}
	\hspace{2cm}

\begin{prooftree}
  \hypo{\vdash \Gamma}
  \infer1[$\wk$]{\vdash \Gamma, A}
\end{prooftree}}

\mbox{
\begin{prooftree}
\infer0[$\ttt$]{\vdash \ttt}
\end{prooftree}
\hspace{2cm}

\begin{prooftree}
  \hypo{\vdash \Gamma}
	\infer1[$\fff$]{\vdash \Gamma, \fff}
\end{prooftree}}

\mbox{
\begin{prooftree}
  \hypo{\vdash \Gamma, A}
  \infer1[$\forall$]{\vdash \Gamma, \forall x A}
\end{prooftree}
($x \notin fv(\Gamma)$)
\hspace{1.2cm}

\begin{prooftree}
  \hypo{\vdash \Gamma, A[t/x]}
  \infer1[$\exists$]{\vdash \Gamma, \exists x A}
\end{prooftree}}

In this paper, we also use some \bfit{deep inference} \cite{BT 2001} rules that are in the
following form:
\begin{center}
\begin{prooftree}
  \hypo{\vdash S\{A\}}
  \infer1[]{\vdash S\{B\}}
\end{prooftree}
\end{center}
where $S\{ \ \}$ stands for a \bfit{context}, which corresponds to a sequent
with a hole taking the place of an atom, and $S\{A\}$ represents
the sequent or formula obtained by replacing the hole in $S\{ \ \}$ with the formula $A$. Formally,

  \[C ::= \Box \ | \ A \cor C \ | \ C \cand A \ | \ \exists x C \ | \ \forall x C \].
  \[S ::= C \ | \ A, S \ | \ S, A \] where $A$ is a formula.
The above rule can be thus seen as the rewriting rule $A \rightarrow B$.

@@

Sequents are proved using the inference rules of $\FOMLL$:
\begin{center}

\mbox{
\begin{prooftree}
  \infer0[$\ax$]{\vdash A, \neg A}
\end{prooftree}
\hspace{2cm}

\begin{prooftree}
  \hypo{\vdash \Gamma, A}
  \hypo{\vdash \Delta, \neg A}
  \infer2[$\cut$]{\vdash \Gamma, \Delta}
\end{prooftree}}

\mbox{
\begin{prooftree}
  \hypo{\vdash \Gamma, A}
  \hypo{\vdash \Delta, B}
  \infer2[$\conj$]{\vdash \Gamma, \Delta, A \cand B}
\end{prooftree}
\hspace{1.5cm}

\begin{prooftree}
  \hypo{\vdash \Gamma, A, B}
  \infer1[$\disj$]{\vdash \Gamma, A \cor B}
\end{prooftree}}

\mbox{
\begin{prooftree}
  \hypo{\vdash \Gamma, A}
  \infer1[$\univ$]{\vdash \Gamma, \forall x A}
\end{prooftree}
($x \notin fv(\Gamma)$)
\hspace{2cm}

\begin{prooftree}
  \hypo{\vdash \Gamma, A[t/x]}
  \infer1[$\exist$]{\vdash \Gamma, \exists x A}
\end{prooftree}}
\end{center}

@@

    \begin{center}	
    \begin{tikzpicture}
    \tikzstyle{every node}=[circle, minimum size=12pt, inner sep=2pt]
    \draw (0, 0) node (par) {$\cor$};
    \draw (1, 1.7) node (dots1) {$\vdots$};
    \draw (-1, 1.7) node (dots2) {$\vdots$};
    \draw (1, 3.4) node (vr) {$v_R$};
    \draw (-1, 3.4) node (vl) {$v_L$};
    \draw (1.5, -1) node (dots3) {$\vdots$};
    \draw (-1.5, -1) node (dots4) {$\vdots$};
    \draw (2.9, -1) node (ul) {$u_L$};
    \draw (-2.9, -1) node (ur) {$u_R$};
    \draw (par) -- (dots1) [color=blue];
    \draw (par) -- (dots2) [color=red];
    \draw (dots1) -- (vr) [color=blue]; 
    \draw (dots2) -- (vl) [color=red];
    \draw (dots4) -- (ur) [color=blue]; 
    \draw (dots3) -- (ul) [color=red];
    \draw (dots4) -- (par) [color=blue]; 
    \draw (dots3) -- (par) [color=red];
    \end{tikzpicture}

    {\bf Figure 7.} A schema showing that the two branches of the same $\cor$ cannot be used in the cycle at the same time. 
    \end{center}

@@

\begin{center}
\begin{tikzpicture}
  \tikzstyle{every node}=[circle, minimum size=20pt, inner sep=2pt]
  \draw (0, 0) node (par1) {$\cor$};
  \draw (1.8, 1.2) node (z) {\color{blue} $\forall z$};
  \draw (-1.8, 1.2) node (par2) {$\cor$};
    \draw (-2.8, 2.4) node (x) {\color{red} $\exists x$};
    \draw (-0.8, 2.4) node (y) {$\exists y$};
    \draw (-2.8, 3.6) node (px) {$\overline{p}x$};
    \draw (-0.8, 3.6) node (qy) {$\overline{q}y$};
    \draw (1.8, 2.4) node (tens) {$\cand$};
    \draw (0.8, 3.6) node (pz) {$pz$};
    \draw (2.8, 3.6) node (qfz) {$qfz$};
    \draw (par1) -- (par2) [<-];
    \draw (par1) -- (z) [<-];
    \draw (par2) -- (x) [<-]; 
    \draw (par2) -- (y) [<-];
    \draw (x) -- (px) [<-]; 
    \draw (y) -- (qy) [<-];
    \draw (z) -- (tens) [<-]; 
    \draw (tens) -- (pz) [<-];
    \draw (tens) -- (qfz) [<-];
    \draw (y) -- (z) [dotted, ->];
    \draw (x) -- (z) [dotted, ->];
    \draw[color=blue, dashed] (px) to [bend left] (pz);
    \draw[color=red, dashed] (qy) to [bend left] (qfz);
\end{tikzpicture}
\vspace{0.5cm}

@@

{\bf Figure 8.} A unification net and its frame. The colored part shows how the dependency ${\color{red} \exists x} \rightarrow {\color{blue} \forall z}$ is transformed.
\end{center}

@@

\begin{center}
\mbox{
\begin{prooftree}
  \hypo{\vdash S\{a \cor a\}}
  \infer1[$\acD$]{\vdash S\{a\}}
\end{prooftree}
\hspace{0.3cm}

\begin{prooftree}
  \hypo{\vdash S\{(A \cand B) \cor (C \cand D)\}}
  \infer1[$\me$]{\vdash S\{(A \cor C) \cand (B \cor D)\}}
\end{prooftree}}
\vspace{0.2cm}

\mbox{
\begin{prooftree}
  \hypo{\vdash S\{\exists x A \cor \exists x B\}}
  \infer1[$\mee$]{\vdash S\{\exists x (A \cor B)\}}
\end{prooftree}	
\hspace{0.3cm}

\begin{prooftree}
  \hypo{\vdash S\{\forall x A \cor \forall x B\}}
  \infer1[$\meee$]{\vdash S\{\forall x (A \cor B)\}}
\end{prooftree}	
}
\end{center}

>>>>

\documentclass[conference,twosided,10pt]{IEEEtran}
% Some Computer Society conferences also require the compsoc mode option,
% but others use the standard conference format.
%
% If IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it like:
% \documentclass[conference]{../sty/IEEEtran}

%%line and pagenumbers to be removed for final version
\usepackage[switch]{lineno}
\pagestyle{plain}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\linenumbers
\section{Introduction}

@@

\bfit{Terms} $s, t, u, \ldots$ and \bfit{atoms} (atomic formulas) $\alpha, \beta,
\gamma, \ldots$ are generated inductively from variables $x, y, z, \ldots$: if
$\delta$ (resp. $p$) is an $n$-ary function (resp. predicate) symbol and $t_1, \ldots,
t_n$ are terms, then $\delta t_1 \ldots t_n$ (resp. $pt_1 \ldots t_n$) is a term
(resp. atom). The set of atoms is extended with logical constant $\ttt$ and
$\fff$. We also assume that each predicate symbol $p$ has a \bfit{dual}
predicate symbol $\overline{p}$ with $\overline{p} \neq p$ and
$\overline{\overline{p}} = p$. We extend the duality to atoms with
$\overline{pt_1 \ldots t_n} = \overline{p}t_1 \ldots t_n$, $\overline{\ttt} =
\fff$, and $\overline{\fff} = \ttt$. \bfit{Formulas} $A, B, C$ are generated inductively from atoms by binary
connectives $\cor$ and $\cand$ and quantifiers $\forall x$ and $\exists x$.
The \bfit{negation} $\neg$ is defined in the usual way. A formula is
\bfit{rectified} if all bound variables are distinct from one another and from
all free variables. Without loss of generality, we assume that all formulas are
rectifed throughtout the paper as every formula has a logically equivalent
rectified form.

@@

A \bfit{sequent} is a multiset of formulas and sequents are proved using the
following inference rules:

\begin{figure}[h]
\begin{center}
\begin{prooftree}
  \infer0[$\ax$]{\vdash A, \neg A}
\end{prooftree}
\qquad
\begin{prooftree}
  \hypo{\vdash \Gamma, A}
  \hypo{\vdash \Delta, \neg A}
  \infer2[$\cut$]{\vdash \Gamma, \Delta}
\end{prooftree}
\\[1.5ex]
\begin{prooftree}
  \hypo{\vdash \Gamma, A}
  \hypo{\vdash \Delta, B}
  \infer2[$\cand$]{\vdash \Gamma, \Delta, A \cand B}
\end{prooftree}
\qquad
\begin{prooftree}
  \hypo{\vdash \Gamma, A, B}
  \infer1[$\cor$]{\vdash \Gamma, A \cor B}
\end{prooftree}
\\[1.5ex]
\begin{prooftree}
  \hypo{\vdash \Gamma, A, A}
  \infer1[$\ctr$]{\vdash \Gamma, A}
\end{prooftree}
\qquad
\begin{prooftree}
  \hypo{\vdash \Gamma}
  \infer1[$\wk$]{\vdash \Gamma, A}
\end{prooftree}
\\[1.5ex]
\begin{prooftree}
\infer0[$\ttt$]{\vdash \ttt}
\end{prooftree}
\qquad
\begin{prooftree}
  \hypo{\vdash \Gamma}
	\infer1[$\fff$]{\vdash \Gamma, \fff}
\end{prooftree}
\\[1.5ex]
\begin{prooftree}
  \hypo{\vdash \Gamma, A}
  \infer1[$\forall$]{\vdash \Gamma, \forall x A}
\end{prooftree}
($x \notin fv(\Gamma)$)
\qquad
\begin{prooftree}
  \hypo{\vdash \Gamma, A[t/x]}
  \infer1[$\exists$]{\vdash \Gamma, \exists x A}
\end{prooftree}
\caption{Gentzen's sequent calculus $\LK$}
\end{center}
\end{figure}

In this paper, we also use some \bfit{deep inference} \cite{BT 2001} rules that are in the
following form:
\begin{center}
\begin{prooftree}
  \hypo{\vdash S\{A\}}
  \infer1[]{\vdash S\{B\}}
\end{prooftree}
\end{center}
where $S\{ \ \}$ stands for a \bfit{context}, which corresponds to a sequent
with a hole taking the place of an atom, and $S\{A\}$ represents
the sequent or formula obtained by replacing the hole in $S\{ \ \}$ with the formula $A$. Formally, 
\begin{center}
  $C ::= \Box \ | \ A \cor C \ | \ C \cand A \ | \ \exists x C \ | \ \forall x 
C$.
\\[1.5ex]
  $S ::= C \ | \ A, S \ | \ S, A$ 
\end{center}
where $A$ is a formula.
The above rule can be thus seen as the rewriting rule $A \rightarrow B$.

@@

Sequents are proved using the inference rules of $\FOMLL$:
\begin{figure}[h]
\begin{center}
\begin{prooftree}
  \infer0[$\ax$]{\vdash A, \neg A}
\end{prooftree}
\qquad
\begin{prooftree}
  \hypo{\vdash \Gamma, A}
  \hypo{\vdash \Delta, \neg A}
  \infer2[$\cut$]{\vdash \Gamma, \Delta}
\end{prooftree}
\\[1.5ex]
\begin{prooftree}
  \hypo{\vdash \Gamma, A}
  \hypo{\vdash \Delta, B}
  \infer2[$\conj$]{\vdash \Gamma, \Delta, A \cand B}
\end{prooftree}
\qquad
\begin{prooftree}
  \hypo{\vdash \Gamma, A, B}
  \infer1[$\disj$]{\vdash \Gamma, A \cor B}
\end{prooftree}
\\[1.5ex]
\begin{prooftree}
  \hypo{\vdash \Gamma, A}
  \infer1[$\univ$]{\vdash \Gamma, \forall x A}
\end{prooftree}
($x \notin fv(\Gamma)$)
\qquad
\begin{prooftree}
  \hypo{\vdash \Gamma, A[t/x]}
  \infer1[$\exist$]{\vdash \Gamma, \exists x A}
\end{prooftree}
\end{center}
\caption{Sequent calculus for $\FOMLL$}
\end{figure}

@@

\begin{figure}[h]
    \begin{center}	
    \begin{tikzpicture}
    \tikzstyle{every node}=[circle, minimum size=12pt, inner sep=2pt]
    \draw (0, 0) node (par) {$\cor$};
    \draw (1, 1.7) node (dots1) {$\vdots$};
    \draw (-1, 1.7) node (dots2) {$\vdots$};
    \draw (1, 3.4) node (vr) {$v_R$};
    \draw (-1, 3.4) node (vl) {$v_L$};
    \draw (1.5, -1) node (dots3) {$\vdots$};
    \draw (-1.5, -1) node (dots4) {$\vdots$};
    \draw (2.9, -1) node (ul) {$u_L$};
    \draw (-2.9, -1) node (ur) {$u_R$};
    \draw (par) -- (dots1) [color=blue];
    \draw (par) -- (dots2) [color=red];
    \draw (dots1) -- (vr) [color=blue]; 
    \draw (dots2) -- (vl) [color=red];
    \draw (dots4) -- (ur) [color=blue]; 
    \draw (dots3) -- (ul) [color=red];
    \draw (dots4) -- (par) [color=blue]; 
    \draw (dots3) -- (par) [color=red];
    \end{tikzpicture}

    \caption{A schema showing that the two branches of the same $\cor$ cannot be
	    used in the cycle at the same time.}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
\begin{tikzpicture}
  \tikzstyle{every node}=[circle, minimum size=20pt, inner sep=2pt]
  \draw (0, 0) node (par1) {$\cor$};
  \draw (1.8, 1.2) node (z) {\color{blue} $\forall z$};
  \draw (-1.8, 1.2) node (par2) {$\cor$};
    \draw (-2.8, 2.4) node (x) {\color{red} $\exists x$};
    \draw (-0.8, 2.4) node (y) {$\exists y$};
    \draw (-2.8, 3.6) node (px) {$\overline{p}x$};
    \draw (-0.8, 3.6) node (qy) {$\overline{q}y$};
    \draw (1.8, 2.4) node (tens) {$\cand$};
    \draw (0.8, 3.6) node (pz) {$pz$};
    \draw (2.8, 3.6) node (qfz) {$qfz$};
    \draw (par1) -- (par2) [<-];
    \draw (par1) -- (z) [<-];
    \draw (par2) -- (x) [<-]; 
    \draw (par2) -- (y) [<-];
    \draw (x) -- (px) [<-]; 
    \draw (y) -- (qy) [<-];
    \draw (z) -- (tens) [<-]; 
    \draw (tens) -- (pz) [<-];
    \draw (tens) -- (qfz) [<-];
    \draw (y) -- (z) [dotted, ->];
    \draw (x) -- (z) [dotted, ->];
    \draw[color=blue, dashed] (px) to [bend left] (pz);
    \draw[color=red, dashed] (qy) to [bend left] (qfz);
\end{tikzpicture}
\vspace{1.5ex}

@@

\caption{A unification net and its frame. The colored part shows how the
dependency ${\color{red} \exists x} \rightarrow {\color{blue} \forall z}$ is
transformed.}
\end{center}
\end{figure}

@@

\begin{center}
\begin{prooftree}
  \hypo{\vdash S\{a \cor a\}}
  \infer1[$\acD$]{\vdash S\{a\}}
\end{prooftree}
\qquad
\begin{prooftree}
  \hypo{\vdash S\{(A \cand B) \cor (C \cand D)\}}
  \infer1[$\me$]{\vdash S\{(A \cor C) \cand (B \cor D)\}}
\end{prooftree}
\\[1.5ex]
\begin{prooftree}
  \hypo{\vdash S\{\exists x A \cor \exists x B\}}
  \infer1[$\mee$]{\vdash S\{\exists x (A \cor B)\}}
\end{prooftree}	
\qquad
\begin{prooftree}
  \hypo{\vdash S\{\forall x A \cor \forall x B\}}
  \infer1[$\meee$]{\vdash S\{\forall x (A \cor B)\}}
\end{prooftree}	
\end{center}
